Effective Dynamic Models of Metabolic Networks

### 2016-03-17 Paper_v1.tex
Working in introduction

Metabolic networks, even for simple organisms such as \textit{E. coli}, contain thousands of intracellular metabolites and reactions that are highly interconnected.
Constraints based models and convex analysis techniques have become the standard approaches to interrogate these complex networks.
Constraints based methods such as flux balance analysis (FBA) \cite{2010_orth_NatBiotech} and convex network decomposition approaches such elementary modes (EMs) [REF] or extreme pathways (EPs) [REF] describe intracellular metabolism using the biochemical stoichiometry, and other constraints such as thermodynamically feasibility [REF]
under pseudo-steady state conditions.
However, elementary mode calculations are computational expensive and infeasible for genome scale networks\cite{2004_lee_varner_ko_ieee}.
FBA is frequently used to study genome-scale networks and often results in good approximations for yields, however it fails to estimate intracellular metabolism due to its underdetermined solution space.
Compared with FBA, MFA has good estimations of intracellular metabolism, but is dependent on experimental data.
Just like FBA, HCM approaches cell metabolism as a resource allocation problem towards a specified objective.
But HCM has multiple pathways to select from (elementary modes) and can use a combination of modes to direct its resources through the cell.
HCM has been shown to have comparable internal flux estimations to MFA and is able to estimate dynamic external fluxes with respect to experimental observations \cite{2008_kim_varner_ramkrishna_BiotechProg}.
DFBA is able to account for dynamic external fluxes but fails to incorporate the regulatory processes involved with HCM without the use of boolean rules\cite{2001_covert_schilling_palsson}.
Despite the advantages of HCM, it is restricted to small networks due to the exponential increase of the number of elementary modes for large networks.
Recent work has been done to reduce the number of elementary modes by lumping modes into groups using complex weighting schemes and experimental data\cite{2010_song_ramkrishna}, however, HCM is still limited by EM decomposition.

>>>>

Biotechnology harnesses the power of metabolism to produce products that benefit society.
Thus, a core challenge in biotechnology is the understanding of intracellular metabolic networks.
Mathematical models of metabolic networks are useful tools to understand and ultimately predict how cells utilize nutrients to produce valuable products.
Genome scale metabolic networks contain thousands of coupled intracellular metabolites and biochemical reactions.
Constraints based techniques have become the standard approach to interrogating these complex networks.
Constraints based methods such as flux balance analysis (FBA) \cite{2010_orth_NatBiotech} and convex network decomposition approaches such elementary modes (EMs) [REF] or extreme pathways (EPs) [REF] describe intracellular metabolism using the biochemical stoichiometry, and other constraints such as thermodynamically feasibility [REF]
under pseudo-steady state conditions.
However, elementary mode calculations are computational expensive and infeasible for genome scale networks\cite{2004_lee_varner_ko_ieee}.
FBA is frequently used to study genome-scale networks and often results in good approximations for yields, however it fails to estimate intracellular metabolism due to its underdetermined solution space.
Compared with FBA, MFA has good estimations of intracellular metabolism, but is dependent on experimental data.
Just like FBA, HCM approaches cell metabolism as a resource allocation problem towards a specified objective.
But HCM has multiple pathways to select from (elementary modes) and can use a combination of modes to direct its resources through the cell.
HCM has been shown to have comparable internal flux estimations to MFA and is able to estimate dynamic external fluxes with respect to experimental observations \cite{2008_kim_varner_ramkrishna_BiotechProg}.
DFBA is able to account for dynamic external fluxes but fails to incorporate the regulatory processes involved with HCM without the use of boolean rules\cite{2001_covert_schilling_palsson}.
Despite the advantages of HCM, it is restricted to small networks due to the exponential increase of the number of elementary modes for large networks.
Recent work has been done to reduce the number of elementary modes by lumping modes into groups using complex weighting schemes and experimental data\cite{2010_song_ramkrishna}, however, HCM is still limited by EM decomposition.
